# Data.frame/table brass tacks (the pointy part)

Matt Dowle and Arun Srinivasan's^[along with many other contributors]
data.table package is a data.frame twin with amazing performance. Often, it's
the fastest way to operate on in-memory tabular data, period (not just in R!).

I've used data.table for performance reasons in projects for many years now.
There are a few things to watch out for outlined in these notes (I have
stumbled across both in practice). Fortunately they are not that problematic to
deal with.


## Data.frames

R's data.frame is simply an R list with extra attributes: class = "data.frame"
and  row.names = (integer or character vector of row names).  Data.frame column
names (if any) are taken from the list names.

That is, you can make a data.frame the usual way:

```{r}
x <- data.frame(v = 1:2, a = c(pi, 5))
```

or, like this if you prefer^[Mostly, there are some edge-case exceptions to
this however!]:

```{r}
y <- structure(
       list(v = 1:2, a = c(pi, 5)),
       class = "data.frame",
       row.names = 1:2)
all.equal(x, y)
```

R defines the number of rows of a data.frame to be the length of its
`row.names` vector attribute. The number of columns is simply the
length of the list.


This simple scheme is very flexible! Each data.frame list entry (column) is effectively
a memory pointer to a native R vector. And usually, R tries to defer copying
data until it must. Consider the following example run on my laptop with
8GB RAM:

```{r}
system.time({
  x <- runif(4e8)
  y <- runif(4e8, 0, 10)
})

system.time({d <- data.frame(y, x)})
```

Each vector x and y consume about 3GB of RAM for a total of about 6GB.
Creation of the data.frame d above happens instantly without data copy because
its columns simply contain pointers to y and x. (Of course, subsequently
modifying the values would induce a copy.)

Re-arranging columns in a data.frame simply re-arranges pointers and need not
always copy data:

```{r}
system.time({print(tail(d[, order(names(d))]))})
```
That operation (ordering columns and printing the tail) ran almost
instantly on my laptop.



## Data.table corner cases

Data.tables are blobs of data behind a reference pointer that (mostly) emulate
standard data.frame structure and behavior.  They are almost always faster and
more efficient than standard R data.frames (and I often encourage their use),
but there exist situations where plain old data.frames can be more efficient.

First, let's run rm and gc to free up the 6GB memory used by the previous example
before proceeding:

```{r}
rm(list = ls())
gc()
```

Now consider this:
```{r}
library(data.table)
system.time({
  dt <- data.table(x = runif(4e8), y = runif(4e8, 0, 10))
})
```
Why did that take so long? Because data.table made copies of the data,
which happen to be larger than the RAM in my laptop in this case, and
thus operating system swap was involved.

Somewhat more problematically, unlike base-R data.frames, simply re-ordering
data.table columns sometimes induces data copies. For example:

```{r}
system.time({print(tail(dt[, order(names(dt)), with = FALSE]))})
```

That's a bit of a bummer--watch out for it. The work around here is usually to
try to use "the data.table way" as much as possible to solve problems.


## Data.table function arguments

Another brass tack with data.table that you should be aware of is that
data.table function arguments are call-by-reference, unlike the standard R
function argument call-by-value convention. (See, for instance,
https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Argument-evaluation).
I've actually personally wrestled with hard-to-find bugs resulting from this, so it is
something to watch out for. To be fair, environment function arguments are also
call-by-reference in R, so it may help to think of data.tables like that.


## Sort order

Be aware that data.table sorts using the 'C' locale (ASCII, basically), which
is almost certainly *not* the way your R session defaults to sorting things.

Here is an example:

```{r}
set.seed(1)
a <- sample(c(letters, LETTERS), 10)

x <- data.table(a = a)
setkeyv(x, "a")  # sort the data.table by the contents of 'a'
a <- sort(a)     # sort 'a' in plain old R and compare:
cbind(x, a)
```

Forcing R to use the C local makes the sort order the same:

```{r}
Sys.setlocale("LC_COLLATE", "C")
a <- sort(a)
cbind(x, a)
```



<br/> <br/> <br/> <br/>
